`make.traits`

### 2024-06-28 inst/gitbook/02_make.traits.Rmd
added vignette for link.traits

>>>>

## Linking traits together with `link.traits`

In the `treats` algorithm, traits are simulated in parallel for improving performances.
Most of the cases this is a (really) good thing!
However, in some cases you might need some of your traits to be simulated sequentially.
For example if you want a specific trait process or options to be dependent on the result of another trait.
You can set up these linked traits using the `link.traits` function.
This function takes `base.trait`, a `"treats"` `"traits"` object created using `make.traits` that will be the first trait simulated and the `next.trait`, a list of  `"treats"` `"traits"` objects that will be simulated after the `base.trait` depending on the link type (`link.type`) and some other arguments (`link.args`).
So far the `link.traits` function only accommodates `link.type = "conditional"` for conditional traits.

### Conditional traits

Say for example that you have a primary trait (`base.trait`) that is a discrete binary trait that generates values `0` or `1`.
This can correspond to some abiotic trait for example (e.g. "is the lineage on an island or on the mainland").
You can first generate this trait using the `make.traits` function as detailed above (note here we're using a transition matrix as a specific argument that makes the changes from state 0 to state 1 relatively rare and the reversal even more rare).

```{r}
## Setting up a discrete trait
discrete_trait <- make.traits(discrete.process,
       process.args = list(transitions = matrix(c(3, 0.2, 0.05, 3), 2, 2)),
       trait.names  = "discrete")
```

Following this trait value, we can create another trait that is conditional to the outcome of this first discrete trait simulation.
In other words we can specify a choice between two processes depending on the result of the `base.trait`.
For example, we can say that if the lineage evolves on an island (state `"1"`) it will evolve using a Brownian motion process but if it evolves on the mainland (state `"0"`), it will evolve using an OU process.

We'll need to first generate these two traits independently using `make.traits` (using different options):

```{r}
## A standard Brownian Motion trait
BM_trait <- make.traits(process = BM.process)
## An OU trait with a strong alpha
OU_trait <- make.traits(process = OU.process, process.args = list(alpha = 10))
## Combining both traits into a list
continuous_trait <- list("mainland_case" = BM_trait,
                         "island_case"   = OU_trait)
```

We then need to create a list of functions to interpret the results of the first trait (`discrete_trait`) so that if the results of `discrete_trait` is `0`, `BM_trait` is used, if it's `1` `OU_trait` is used.
This is done by creating a list of functions that evaluate `x1`, the outcome of `discrete_trait`, and return a logical (`TRUE` or `FALSE`).
For example `function(x1) {x1 != 0}` will return `TRUE` if the `discrete_trait` is not state `0`:

```{r}
## Creating a list of functions to choose between either case for the continuous trait
select_continuous <- list("is.mainland" = function(x1) {x1 == 0},
                          "is.island"   = function(x1) {x1 == 1})
```

Note that this list must be in the same order as the `continuous_trait`: the functions will be evaluated sequentially (i.e. test the function `is.mainland`, then `is.island`) and the first one to return `TRUE` will be used to select the function in `continuous_trait`.
For example in our case if the score of `discrete_trait` is `1`, the second function from `select_continuous` is `TRUE` so the second element from `continuous_trait` is used (the `OU_trait`).

We can then combine all that using `link.traits`:

```{r}
## Link these traits
conditional_trait <- link.traits(# The first trait to be evaluated
                                 base.trait = discrete_trait,
                                 # The other traits to be evaluated next
                                 # (here a list of two traits)
                                 next.trait = continuous_trait,
                                 link.type  = "conditional",
                                 # The functions to choose between either
                                 # continuous traits 
                                 link.args  = select_continuous)
```

And run the usual simulation pipeline with `treats`:

```{r, fig.height = 6, fig.width = 12}
## Running a tree with this conditional trait
set.seed(123)
my_conditional_tree <- treats(stop.rule = list(max.taxa = 200),
                              traits    = conditional_trait)

op <- par(mfrow = c(1,2))
plot(my_conditional_tree, trait = 1, main = "The conditional trait")
plot(my_conditional_tree, trait = 2, main = "The conditioned trait")
par(op)
## The clumpy bit around the value 0 corresponds to the OU trait
```

Of course, you can combine linked traits with other traits like any other traits in `treats` using `make.traits` (for example: `make.traits(process = my_new_trait, add = conditional_trait)` to generate more complex traits.
In this case the traits are generated in parallel (for speed) apart for the specified linked ones.
